---
title: "Introduction to CineGrade"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CineGrade}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  eval       = FALSE   # All chunks eval=FALSE for CRAN/R CMD CHECK safety
)
library(CineGrade)

# Convenience: paths to built-in sample images
portrait <- system.file("extdata", "sample_portrait.jpg", package = "CineGrade")
wes      <- system.file("extdata", "preset_wes.jpg",      package = "CineGrade")
noir     <- system.file("extdata", "preset_noir.jpg",     package = "CineGrade")
vintage  <- system.file("extdata", "preset_vintage.jpg",  package = "CineGrade")
anime    <- system.file("extdata", "preset_anime.jpg",    package = "CineGrade")
```

## What is CineGrade?

**CineGrade** extracts the cinematic colour grade from any reference image and
mathematically transfers it to a target photo — making your photo look like a
scene from a Wes Anderson film, a neon-noir thriller, or any style you choose.

Every transformation is powered by a technique from the MTH 209 / MTH 210
curriculum:

| Technique | Function(s) |
|:---|:---|
| **k-means clustering** in LAB colour space | `extract_grade()`, `apply_grade()`, `stylize()` |
| **1-nearest-neighbour matching** (kNN) | `apply_grade()` centroid matching step |
| **Principal Component Analysis** (PCA) | `palette_pca()` |
| **UMAP** dimensionality reduction | `color_universe()` |
| **Euclidean distance** in LAB (ΔE) | `compare_grades()` |
| **Matrix transformation** (CVD) | `simulate_colorblind()` |

The package also ships a full interactive **Shiny dashboard** (`run_cinegrader()`)
for real-time before/after preview and download.

---

## Why LAB colour space?

All core algorithms run in **CIE LAB** (L\*, a\*, b\*) rather than RGB because:

- Euclidean distance in LAB ≈ human-perceived colour difference
  (ΔE = 1 ≈ one just-noticeable difference)
- k-means minimises within-cluster sum of squares — in LAB this corresponds
  directly to *perceptual* homogeneity, not just numeric similarity

The two utility functions `rgb_to_lab()` and `lab_to_rgb()` handle all
conversions internally via the `colorspace` package.

---

## 1. Extracting a Colour Grade — `extract_grade()`

`extract_grade()` runs k-means on a random sample of pixels in LAB space and
returns the **k cluster centres** as the image's compact colour signature.

```{r extract}
grade <- extract_grade(wes, k = 8)
print(grade)
```

The `CineGrade` S3 object carries:

- `$centers` — k × 3 matrix of LAB centroids (sorted by L* in `palette_pca`)
- `$labels`  — integer vector of per-pixel cluster assignments
- `$k`       — the k used
- `$colorspace` — always `"LAB"`

**Why k-means?** Each row of `$centers` is a "prototype" colour that
summarises a region of the image's colour palette. With k = 8, the entire
colour story of a film still is compressed to 8 numbers in 3D LAB space.

---

## 2. Transferring a Colour Grade — `apply_grade()`

`apply_grade()` is where the magic happens. It implements a four-step
**centroid-shift algorithm**:

1. Extract k reference centroids (from the reference image).
2. Extract k target centroids **and** per-pixel labels (from the target).
3. **1-NN match**: for each of the k target centroids, find the nearest
   reference centroid in LAB Euclidean space.
4. **Pixel shift**: for every pixel assigned to target cluster *j*, add the
   delta Δ_j = ref_centroid[match(j)] − target_centroid[j] to its LAB values.

```{r apply}
result <- apply_grade(portrait, wes, k = 8)

# Display side by side
par(mfrow = c(1, 2), mar = c(0, 0, 1.5, 0))
plot(CineGrade:::load_image(portrait), axes = FALSE)
title("Original", col.main = "grey40")
plot(result, axes = FALSE)
title("Wes Anderson grade", col.main = "grey40")
par(mfrow = c(1, 1))
```

You can also feed a **pre-computed `CineGrade` object** as the reference to
avoid re-running k-means:

```{r apply_precomputed, eval = FALSE}
wes_grade <- extract_grade(wes, k = 8)
result2   <- apply_grade(portrait, wes_grade, k = 8)
```

---

## 3. Measuring Style Distance — `compare_grades()`

`compare_grades()` quantifies how different two images' colour styles are,
using the **symmetric average Hausdorff distance** in LAB space:

$$d(G_1, G_2) = \frac{1}{2}\left[\frac{1}{k}\sum_i \min_j\|c^1_i - c^2_j\|_2 + \frac{1}{k}\sum_j \min_i\|c^1_i - c^2_j\|_2\right]$$

The result is in **ΔE units** — the same perceptual scale as CIEDE:

| Score | Meaning |
|:---|:---|
| 0 | Identical colour grade |
| < 5 | Very similar (subtle difference) |
| 5 – 15 | Noticeable but related styles |
| 15 – 30 | Clearly different styles |
| > 30 | Strongly different (e.g. noir vs. anime) |

```{r compare}
d_wes_noir <- compare_grades(wes, noir,    k = 8)
d_wes_vint <- compare_grades(wes, vintage, k = 8)

cat("Wes vs Noir:    ΔE =", round(d_wes_noir, 2), "\n")
cat("Wes vs Vintage: ΔE =", round(d_wes_vint, 2), "\n")

print(d_wes_noir)
```

---

## 4. Visualising the Colour Universe — `color_universe()`

`color_universe()` takes a random sample of pixels, embeds their LAB
coordinates in 2D using **UMAP**, and renders each pixel as a coloured dot at
its 2D position.  The result is a visual **colour fingerprint**:

```{r universe, fig.height = 5}
p_wes  <- color_universe(wes,  n_pixels = 3000, title = "Wes Anderson",  seed = 1)
p_noir <- color_universe(noir, n_pixels = 3000, title = "Noir",          seed = 1)
print(p_wes)
```

- Sunset/warm images → tight warm orange-yellow clouds
- Noir → dense dark scatter near the bottom-left
- Anime → wide, high-chroma scatter with vivid outliers

---

## 5. Comparing Styles Across Images — `palette_pca()`

`palette_pca()` builds a feature matrix where each image is represented by
its **L*-sorted k-means centroids** (flattened to a 3k-dimensional vector),
then runs **PCA** to project all images into a 2D style space.

**Why sort centroids?**  k-means label ordering is arbitrary — centroid 1
from image A is not comparable to centroid 1 from image B.  Sorting by L*
(dark → bright) creates a canonical ordering so that the PCA features are
meaningfully comparable across images.

```{r pca, fig.height = 5}
images <- list(
  "Wes Anderson" = wes,
  "Noir"         = noir,
  "Vintage"      = vintage,
  "Anime"        = anime,
  "Portrait"     = portrait
)
p <- palette_pca(images, k = 6)
print(p)
```

Each point is filled with the image's **dominant (highest-L\*)** cluster
colour, labelled with the image name.  The axis labels show the proportion of
variance explained by each principal component.

---

## 6. Artistic Stylization — `stylize()`

`stylize()` applies five artistic modes, all based on k-means quantization:

```{r stylize, fig.height = 4}
poster <- stylize(portrait, mode = "poster", k = 8)
noir_s <- stylize(portrait, mode = "noir")
anime_s<- stylize(portrait, mode = "anime",  sat_scale = 2.0)
pixel_s<- stylize(portrait, mode = "pixel",  k = 6, block_size = 16)

par(mfrow = c(2, 2), mar = c(0, 0, 1.5, 0))
plot(poster,  axes = FALSE); title("poster",       col.main = "grey30")
plot(noir_s,  axes = FALSE); title("noir",         col.main = "grey30")
plot(anime_s, axes = FALSE); title("anime",        col.main = "grey30")
plot(pixel_s, axes = FALSE); title("pixel (16px)", col.main = "grey30")
par(mfrow = c(1, 1))
```

| Mode | Technique | Effect |
|:---|:---|:---|
| `"poster"` | k-means LAB snap | Flat graphic / screen-print |
| `"pixel"` | Downsample → snap → upsample | Retro pixel-art |
| `"anime"` | Chroma-boosted LAB snap | Bold high-saturation colours |
| `"noir"` | Desaturate + 4-cluster snap | Black-and-white with cool tint |
| `"vintage"` | `apply_grade()` + built-in preset | Warm faded film look |

---

## 7. Colour Vision Deficiency Simulation — `simulate_colorblind()`

`simulate_colorblind()` applies standard **Machado et al. (2009)** CVD
transformation matrices in linear sRGB space:

$$C_{\text{out}} = \left(\text{clamp}\left(C_{\text{in}}^{2.2} \cdot M_{\text{CVD}}^\top\right)\right)^{1/2.2}$$

```{r colorblind, fig.height = 3}
deut <- simulate_colorblind(portrait, type = "deuteranopia")
prot <- simulate_colorblind(portrait, type = "protanopia")

par(mfrow = c(1, 3), mar = c(0, 0, 1.5, 0))
plot(CineGrade:::load_image(portrait), axes = FALSE); title("Original")
plot(deut, axes = FALSE); title("Deuteranopia (green-blind)")
plot(prot, axes = FALSE); title("Protanopia (red-blind)")
par(mfrow = c(1, 1))
```

Use this to check whether your colour-graded output remains interpretable for
viewers with impaired colour vision — an important **accessibility check**
before sharing any colour-critical work.

---

## 8. Interactive Dashboard — `run_cinegrader()`

```{r shiny, eval = FALSE}
run_cinegrader()
```

The Shiny app provides:

- **Target image upload** (JPG/PNG)
- **Reference**: built-in film presets (Wes Anderson, Noir, Vintage, Anime)
  **or** upload your own
- **k slider** (3–20)
- **Before / After split-screen** with live ΔE score
- **Download** the graded image as JPEG

---

## Session Info

```{r session, eval = TRUE}
sessionInfo()
```
